draft 1

### 2016-07-03 finalwriteup/draft.rmd
for duncan

# Glossary

ANOVA - Anaylis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.
B2B - Business to Business 
CBR - Case based reasoning. A method of estimating by using the results of similar cases.
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.
DAG - directed acyclic graph. A graph consisting of vertices connected by edges, where each edge directs a vertex to another vertex. The arrangement of the edges is such that a vertex cannot follow a path that loops back to itself.
IT - Information technology
BIM - Building Information Modelling. A digital representation of physical characteristics of an object or structure.
ROC - Reciever Operating Characteristic
AUC - Area under the curve. In this project, the area refers to the area under an ROC curve
TPR - True positive rate
FPR - False positive rate
TP - true positive
FP - false positive
DA - Discriminant analysis
CLV - Customer Lifetime Value
IQR - Interquartile range
ESS - error sum of squares
RMSE - root mean squared error
FWLS - feature weighted linear stacking

@@

# Introduction - 2 pages

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of these sort every day- guessing how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the individual is attempting the task for the first time. These errors are trivial in everyday activities, but they have more meaning for businesses that sell their time as their chief revenue. Consulting businesses give expert advice to other professionals, in exchange for a fee for the amount of time a constultant spent on the problem. The delivery of 'expert advice' is generally in the form of a document (which could be anything from a report to hundreds of drawings) and can range from an hours advice to thousands of hours work on a substantial project. Before a project commences, the client and consultant must agree on a fee for the expert advice, or at least a fee structure. The nature of how a fee is formulated can be as creative as the engaged parties wish, however a couple of simple examples are as follows: the client agrees to pay for the consultants time by the hour until the task is complete, or the consultant provides a fixed fee to complete the project in full, regardless of their hours spent. Both contracts assign financial risk to one party which they must then mitigate to the best of their abilities.

This thesis focuses on the risk taken upon by a consulting company when agreeing to a fixed fee. Clients commonly collect fixed quotes from several consultants before settling with one or negotiating further so fees must be competitive. The way a consulting manager calculates their fixed fee varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project he is quoting and can use a combination of intuition, rules of thumb (such as a fraction of the entire project cost), and comparison to past projects. Such methods can result in a range of profit margins which ultimately depend on how well the manager predicts the amount of time a project demands. This thesis examines the problem of cost prediction via a case study on a company's internal project performance data for the previous twelve years. The study tests whether a statistical model based on historical project data can predict the profitability of a project using characteristics available at the beginning of a project. Furthermore, this thesis investigates how the model's predictions could impact a consulting manager's determination of a fixed fee and the resulting impact on the company's bottom line.

# Problem Description

@@

## Research Motivation - 4.5 pages

@@

Project managers have struggled with forecasting project costs accurately for decades, across many industries. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget on average [@Flyvbjerg2011]. These statistics reveal a concerning glimpse at projects' hidden financial stories which can cause job losses, ruin private businesses, and churn through government balancesheets. Complex long term projects are necessary for modern society producing infrastructure, houses, websites, and movies. This research focuses on the companies who complete portions of these complex projects and the financial risk that is contractually transferred to them. Specifically, the risk of the project owner is transferred to their sub-consultant by mandating fixed flat fee structures to complete a given amount of work. The subconsultant commits to completing their role in a project for a nominated price, regardless of whether the work takes less or more time than intially anticipated. In a similar way to infrastructure and IT project budgets, consultants are susceptible to underestimating the amount of time a project will demand, and eventually wear the cost of extra employee hours spent. For privately owned businesses, mitigating this risk would increase the chances of the business surviving as well as growing and succeeding.

The ramifications of unprofitable projects are multifold. Employee morale deteoriates and employees may view themselves as incompetent if they are responsible for the project's delivery. Employees also experience stress attempting to complete the project within a disappearing budget and may produce lower quality work in exchange for speed. This experience is unfortunate if in reality, the project fee was under-estimated. Unprofitable projects also limit a business' opportunity to invest money into marketing or staff training, which would foster growth and futher profits. The financial risk taken on fixed flat fee projects may also discourage business' from taking other, more thought-out risks, such as expanding the business into a new area. Logically, predicting fixed fee project costs more accurately could positively affect business growth opportunities and staff morale.

@@

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another, and similar work may demand significantly different amounts of time. For example, an ad filmed in a remote location will cost significantly more than an ad filmed on an existing set, even though the effect of the final advert could be identical. Therefore, a constant fixed price to produce an advert is clearly inappropriate for film projects, but some kind of fixed price is often demanded by the client before a project is awarded. Charging clients by the hour is not competitive enough in many industries. Other consultancies are often willing to offer a fixed flat fee for the same work, luring the client with reduced financial risk on their part.

Another strategy to mitigate risks could be to track employee hours, and stop work once the fixed consulting fee has been depleted. Then, negotiate further fees, or variations, with the client and persuade them to accept. This satisfactorily reduces consultant financial risk, and in some cases, depending on the nature of the work or uncontrollable changes, this method can be effective. However, if the project has been following a predictable process but the hours are nevertheless inflating beyond expectation (a common situation), consultants may prefer to wear losses or endure marginable profits to preserve the relationship with their client. Consultants may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction. Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed fees in complex projects. In many industries, the reality is fixed flat fee structures are expected and business' must conform to survive.

@@

Limited research is available on complex-project cost estimation techniques performed by industry. However, two surveys are available from 2000: one on construction projects, the other on IT projects. The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practises. They found that the main method for cost estimation was breaking the project into detailed parts and adding up the cost of each item. The next two most popular methods were 'comparison with similar projects based on documented facts', and 'comparison with similar projects based on personal experience'. These can all be classified as experience based models [@Akintoye2000].

A survey by Moores and Edwards in 1992 of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem. Although limited recent literature is available on industry cost estimation practises, historically it can be observed that for complex construction and software projects, cost predictions are chiefly made with a mixture of experience, project comparisons, and an analysis of the project details and tasks. It is also acknowledged that cost estimation is a problem and hypothesised that inexperience and lack of time are chief contributors [@Akintoye2000].

@@

@Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. This constrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Locallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project are ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model which is then used to adjust the forecasted cost made by detailed analysis (the inside view) [@Flyvbjerg2011]. Managers rarely analyse data from similar projects even though it could significantly improve cost predictions [@Flyvbjerg2011]. This tendency led to a formalised introduction of 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011].

![This method was implemented by the British government for a rail project which is still under completion](images/lit_review/UKrail.JPG)

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however access to credible data to a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the theory of optimism bias is a credible explanation for the problem of persistant underestimation of project costs. The accompanying solution of reference class forecasting presents a promising solution given enough accurate data.

The theory of reference class forecasting was first definitively published in 2003, however the idea of predicting project costs from previous data has been the subject of research for many preceding years and continues to this day. Although research results have often been promising, industry uptake of the idea is not yet established.

@@

A substantial amount of research on project cost prediction has been dedicated to the fields of software development and construction projects, along with some work in the film industry. These analyses have generally used project information from 15 to 20 large projects. A single dataset may consist of projects completed in different countries with different clients and project teams. The mathematical methods used to predict project costs vary from statistical regressions to deep-learning neural networks [@Love2005][@Kim2004]. Most studies report their predictive models perform well and that generally neural networks are more accurate than linear regression.

@@

Statistical research projects on reference forecasting have shown promising results in IT, construction, and film projects. However, persuading industry to adopt these methods into everyday business is a separate challenge.

@@

A survey of the construction industry in the UK in 2000 found that a manager's experience and intuition is still the dominant method for construction cost calculation [@Akintoye2000]. Surveys on current industry practise in cost estimation are scarce, however another study on the IT industry by @Jorgensen2007 argued that previously developed parametric models were not comprehensive enough to be used in industry and that expert judgment was not mobilised enough in the process of developing these tools.

@@

A Customer Relationship Management software package (CRM) was employed by this company to collect and store project data. It is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, information is recorded against each project as well as each client, client contact, and each invoice.

@@

The case study provided a framework to test the value of a consulting company's CRM data in assisting project cost estimates. However, it was important to be aware of limitations when reviewing outcomes and framing the problem. The most obvious limitation is that a single company is being disected in a specific industry, which limits the ability to state conclusions applicable to a broad range of cases. This can be overcome with future studies. Other constraining factors include various forms of missing data and human data entry error.

@@

The project outcomes are intended to assist the business in  predicting project costs before project engagement. Several statistical and machine learning techniques will be tested, compared and refined.

@@

This work contributes to the body of research in numerous ways. There is a gap in the literature of applying statistical and machine learning methods to internally generated business data to facilitate internal business decisions. Studies have modelled the commercial performance of movies, customer churn, the share market, and bankruptcies but few have modelled more day-to-day data that a business generates which could actually assist day-to-day decisions. Improving business decisions would reduce risk and improve profits in scenarios that could be applicable in thousands of businesses around the world.

In the case of this study, the problem of cost estimation is tackled. This issue has been worked on extensively in past research, however the problem has rarely been addressed using a single company's internal data. In the literature, dozens of projects from other companys and even countries are collected and analysed. This thesis uses the internal CRM data from a company which contains over 2,000 past job cases - far more than the usual cost estimation study. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry that estimates and must stand by their fixed fees before a project is won. This differs from other studies estimating actual construction costs of projects and is likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset and different industry sector, ensemble tree methods will be applied which have been minimally tested on this problem. To date, linear regression, neural networks, case based reasoning, and SVM's have been used to predict cost/effort even though ensemble tree methods can perform as well as neural networks in some cases and generally outperform linear regression. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots provide insight into the model's calculations. This is in contrast to neural networks and support vector machines (SVM's) which are 'black-box' predictors. Providing insight to a user is a major benefit in this case as past research has demonstrated issues with user trust and uptake of models, even if the models were successful. It is anticipated that using internal data only also improves user trust and creates tangible meaning for users. Blending multiple machine learning and statistical models was trialled which has not yet been applied to the cost estimation problem.

@@

The following chapter, Chapter 4, describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios and further applications were considered.

@@

* Chapter 5: Variable Selection
* Chapter 6: Regression Models
* Chapter 7: Binary Classification Models
* Chapter 8: Extended Research including model blending, overall profit analysis, research limitations and future work

The results section in each chapter presents key charts and tables that progressively answer the research question: can a statistical or machine learning model based on a company's internal historical project data predict the profitability of a new project and have a positive impact on the bottom line of the business? Then, a discussion finalises each chapter, where the results are examined deeply and analysed in a broader context. The nature of the results are also debated in terms of how surprising they were, their impact on the research hypothesis and limitations as well as suggested future work. The final chapter, the Conclusion, summarises the findings of the project, how the research answered the hypotheses and how the work could be viably applied in industry.

# Literature Review - 25 pages for pHd

@@

**Detailed Analysis**

@@

**Statistical and Machine Learning Models**

@@

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. They found that artificial neural networks and SVM's were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, neural networks and SVM's recieved a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@Shin2015 pioneered the application of boosted trees to cost estimation in construction projects. This is surprisingly late given the rapid uptake of boosted trees over the past decade. In the study, data from 234 school building construction projects in Korea were used. Boosted trees were compared to neural networks, which are one of the most popular algorithms to apply to this problem. Boosted trees were found to predict costs slightly more accurately than neural networks, but not statstically significantly (p value < 0.5). Even though the two algorithms performed similarly, the output from boosted tree models provide insight into the structure of the model such as variable importance and variable partial dependence plots. This additional insight may mean boosted trees are preferable over neural networks because of their potential to engage decision makers.

A notable amount of literature has studied the predictive power of multiple linear regression in the construction cost estimation problem. Often, linear regression is the only model assessed, without comparison to other methods such as neural networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on linear regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Out of those that have compared regression to neural networks, the studies have shown that neural networks outperform regression, however other studies established they are approximately equal [@Kim2004][@Attalla2003]. @Dissanayaka1999 found that neural networks outperformed multiple linear regression models, however regression models could be performed first to determine variable importance and condense the number of variables. It can be argued that boosted trees could perform as well as neural networks while providing additional insights (even more so than regression models) such as variable importance and partial dependence of variables against the independent variable, cost.

Despite neural networks typically outperforming linear regression models, it is interesting to note some of the outcomes of regression studies in cost estimation. Different studies tended to produce a wide range of variables that contribute at fluctuating levels in the models. This is likely because construction projects are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which contribute varying levels of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs. private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionaires, the available variables also largely depended on the questions in the questionaires, and therefore each study produced a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally reported their mean absolute error or a similar metric from the tested models. Similar result structures are useful, however there is a gap in the literature in how to translate the improved errors from the model into a well justified argument for industry to reduce losses by integrating a parametric model into their business.

@@

**Case Based Reasoning**

@@

**Summary**

>>>>

# Acronyms

**ANOVA** - Anaylis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.

**B2B** - Business to Business. Commerce transactions between two businesses.

**CBR** - Case based reasoning. A method of estimating by using the results of similar cases.

**CRM** - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.

**DAG** - Directed acyclic graph. A graph consisting of vertices connected by edges, where each edge directs a vertex to another vertex. The arrangement of the edges is such that a vertex cannot follow a path that loops back to itself.

**IT** - Information technology. The use of computers for storing, retrieving and processing information.

**BIM** - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

**ROC** - Reciever Operating Characteristic. An ROC curve plots the performance of a binary classification algorithm its scoring threshold is varied.

**AUC** - Area under the curve. In this project, the area refers to the area under an ROC curve. Areas greater than 0.5 indicate predictions performing better than random chance while an area of 1 indicates perfect predictions.

**TP** - True positive. The case where a case marked 'positive' in a dataset is correctly predicted 'positive' by an algorithm. Generally, 'positive' is assigned to categories for which the use case needs to be alerted.

**FP** - False positive. A case marked 'negative' in a dataset is incorrectly predicted 'positive' by an algorithm.

**TN** - True negative. A case marked 'negative' in a dataset is correctly predicted 'negative' by an algorithm.

**FN** - False negative. A case marked 'positive' in a dataset is incorrectly predicted 'negative' by an algorithm.

**TPR** - True positive rate. The proportion of positives that were correctly identified (true positives).

**FPR** - False positive rate. The proportion of positives that were incorrectly identified (false positives).

**DA** - Discriminant analysis. Statistical analysis to predict a categorical response variable.

**CLV** - Customer Lifetime Value. Prediction of the profit a customer will bring over the course of a business' relationship with the customer.

**IQR** - Interquartile range. A measurement of spread or variability for a continuous variable. When the data is ordered numerically and divided into four equally sized portions, the values that divide the portions are called quartiles (first, second, third and fourth quartiles). The first quartile has the highest value of the lowest quarter. The interquartile range is the third quartile value minus the first quartile.

**ESS** - Error sum of squares. The sum of squared errors from a set of predictions.

**RMSE** - Root mean squared error. The square root of the mean of the ESS.

**FWLS** - Feature weighted linear stacking. A linear combination of model results which are interacted with original variables. This allows certain models to be weighted heavier for certain values of the original features.

@@

# Chapter 1: Introduction

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the individual is attempting the task for the first time. These errors are trivial in everyday activities, but they have more meaning for businesses that sell their time as their chief revenue.

Consulting businesses give expert advice to other professionals, in exchange for a fee for the amount of time a consultant spent on the problem. The delivery of 'expert advice' is generally in the form of a document (which could be anything from a report to hundreds of drawings) and can range from an hours advice to thousands of hours work. Before a project commences, the client and consultant must agree on a fee for the expert advice, or at least a fee structure. The nature of how a fee is formulated can be as creative as the engaged parties wish, however a couple of simple examples are as follows: the client agrees to pay for the consultants time by the hour until the task is complete, or the consultant provides a fixed fee to complete the project in full, regardless of their hours spent. Both contracts assign financial risk to one party which must then mitigate the risk to ensure profitability.

This thesis focuses on the risk taken by a consulting company when agreeing to a fixed fee. Clients commonly collect fixed quotes from several consultants before settling with one or negotiating further so fees must be competitive. The way a consulting manager calculates their fixed fee varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and can use a combination of intuition, rules of thumb (such as a fraction of the entire project cost), and comparison to past projects. Such methods can result in a range of profit margins which ultimately depend on how well the manager predicts the amount of time a project demands.

This thesis examines the problem of cost prediction via a case study on a company's internal project performance data for the previous twelve years. Statistical models based on historical project data were tested for their ability to predict profitability of projects using characteristics available at the beginning of a project. Furthermore, the potential for a model's predictions to improve a company's bottom line is investigated. This could be achieved with a strategy for how the model's predictions should change managers' fee proposals for risky projects. Due to the commercially sensitive nature of the data used, the source is not disclosed in this dissertation. Variables have also been renamed to ensure confidentiality is maintained, however, no such obfuscation has been applied in the methods used – which are presented with integrity.

# Chapter 2: Problem Description

@@

## Research Motivation

@@

Project managers have struggled with forecasting project costs accurately for decades, across many industries. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget on average [@Flyvbjerg2011]. These statistics cast a concerning light on projects' hidden financial stories which can cause job losses, ruin private businesses, and shrink government balance sheets.

Although complex projects are more likely to run over budget, they reflect the complexities and priveleges of modern society - from major infrastructure, to housing, websites, and movies. All have functional and/or statutory requirements which are defined by a client, and are discharged through contracts to companies via a financially competitive process. Consulting companies who complete portions of these complex projects accept the financial risk that is contractually transferred to them. More specifically, the risk of the project owner is often transferred to their sub-consultant by mandating fixed flat fee structures to complete a given amount of work. The subconsultant commits to completing their role in a project for a nominated price, regardless of whether the work takes less or more time than initially anticipated. For privately owned businesses, mitigating this risk would increase the chances of the business surviving as well as growing and succeeding.

The ramifications of unprofitable projects are multifold. Employee morale deteriorates and employees may view themselves incompetent if they are responsible for the project's delivery. Employees also experience elevated stress attempting to complete underestimated projects within a disappearing budget and may compromise quality of work in exchange for speed. This experience is unfortunate if in reality, the project fee was under-estimated. Unprofitable projects also limit a business' ability to invest money into marketing, training, and business development which would foster growth. The financial risk taken on fixed flat fee projects may also discourage businesses from taking other, more calculated risks, such as expanding the business into a new area. Logically, improving the prediction of project profitability could positively affect business growth opportunities and staff morale.

@@

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another - similar work may demand significantly different amounts of time. For example, an ad filmed in a remote location will cost significantly more than an ad filmed on an existing set, even though the effect of the final advertisement could be identical. Therefore, a constant fixed price to produce an advert is clearly inappropriate, but some kind of fixed price is often demanded by the client before a project is awarded. Charging clients by the hour is not competitive enough in many industries. Other consultancies are often willing to offer a fixed flat fee for the same work, luring the client with reduced financial risk on their part.

Another strategy to mitigate risk is to agree to track employee hours, and stop work once the fixed consulting fee has been depleted. Then, negotiate further fees, or variations, with the client. This approach may satisfactorily reduce a consultant’s financial risk, however, may cause friction between the client and the consultant and affect potential repeat custom. Especially if the project followed a predictable process but the hours inflated beyond expectation (a common situation), consultants may prefer to wear losses or endure marginable profits to preserve their client relationship. Consultants may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction. Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed fees in complex projects. In many industries, the reality of fixed flat fee structures are unavoidable businesses conform to survive.

@@

Limited research is available on complex-project cost estimation techniques performed by industry. However, two surveys are available from 2000: one on construction projects, the other on IT projects. The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practices. They found that the main method for cost estimation was breaking the project into detailed parts and adding up the cost of each item. The next two most popular methods were 'comparison with similar projects based on documented facts', and 'comparison with similar projects based on personal experience'. These can all be classified as experience based models [@Akintoye2000].

A survey by Moores and Edwards in 1992 of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem. Although recent literature is limited,  cost estimation was identified as a problem and it was hypothesised that inexperience and lack of time were chief contributors [@Akintoye2000].

@@

@Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. This constrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project are ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the inside view) [@Flyvbjerg2011]. Managers rarely analyse data from similar projects even though it could significantly improve cost predictions [@Flyvbjerg2011]. This tendency led to a formalised introduction of 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011]. An example reading from this chart would be "If the chance of a cost overrun of 20% is acceptable, the estimated cost should be increased by 55% from original calculations.". This builds past experiences of cost overruns into the initial cost estimation.

![This model was implemented by the British government for a rail project that is still under completion](images/lit_review/UKrail.JPG)

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however, access to credible data for a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the theory of optimism bias is a credible explanation for the problem of persistant underestimation of project costs. The accompanying solution of reference class forecasting presents a promising solution given enough accurate data.

The theory of reference class forecasting was first definitively published in 2003, however, the idea of predicting project costs from previous data has been the subject of research for many preceding years and continues to this day. Although research results have often been promising, industry uptake of the idea is not yet established.

@@

A substantial amount of research on project cost prediction has been dedicated to the fields of software development and construction projects, along with some work in the film industry. These analyses have generally used project information from 15 to 20 large projects. A single dataset may consist of projects completed in different countries with different clients and project teams. The mathematical methods used to predict project costs vary from statistical regressions to deep-learning Neural Networks [@Love2005][@Kim2004]. Most studies report their predictive models perform well and that generally Neural Networks are more accurate than Linear Regression.

@@

The existing models have shown predictive success, however the data and intention from this case study differs from previous studies. Prediction of construction cost is a different problem from consulting effort estimation and uses different variables (except the variables describing the construction project since the consulting company works in the construction industry). Software effort estimation projects are estimating time, however past studies have compared projects across many companies (obtained by surveys etc) while the case study company has access to their data only. Additionally, their data contains a rich variety of variables which cannot be exctracted from surveys. The use of their own data would also improve trust in model outputs.

@@

A survey of the construction industry in the UK in 2000 found that a manager's experience and intuition is still the dominant method for construction cost calculation [@Akintoye2000]. Surveys on current industry practice in cost estimation are scarce, however, another study on the IT industry by @Jorgensen2007 argued that previously developed parametric models were not comprehensive enough to be used in industry and that expert judgment was not mobilised enough in the process of developing these tools.

@@

A Customer Relationship Management software package (CRM) was employed by this company to collect and store project data. A CRM is a popular type of software used by businesses to record client and project details as well as employee timesheet records. In the case study business, the CRM is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, information is recorded against each project as well as each client, client contact, and each invoice.

@@

The case study provided a framework to test the value of a consulting company's CRM data in assisting project cost estimates. However, it was important to be aware of limitations when reviewing outcomes and framing the problem. The most obvious limitation is that a single company is being disected in a specific industry, which limits the ability to state conclusions applicable to a broad range of companies. This can be overcome with future studies, once a range of optimal methods and variables have been determined. Other constraining factors include various forms of missing data and human data entry error.

@@

The project outcomes are intended to assist the business in predicting project profitability before engagement. Several statistical and machine learning techniques were tested, compared and refined.

@@

Past research has tackled cost estimation in various industries, from different angles and with a range of data, however the problem has rarely been addressed using a single company's internal data. In the literature, dozens of projects from other companys and even countries are collected and analysed. This thesis uses the internal CRM data from a company which contains over 2,000 past job cases - far more than the usual cost estimation study. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry that estimates and must stand by their fixed fees before a project is won. This differs from other studies estimating actual construction costs of projects and is likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset and different industry sector, ensemble tree methods will be applied which have been minimally tested on this problem. To date, Linear Regression, Neural Networks, case based reasoning, and SVM's have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots provide insight into the model's calculations. This is in contrast to Neural Networks and support vector machines (SVM's) which are 'black-box' predictors. Providing insight to a user is a major benefit in this case as past research has demonstrated issues with user trust and uptake of models, even if the models were successful. It is anticipated that using internal data only also improves user trust and creates tangible meaning for users. Blending multiple machine learning and statistical models was trialled which has not yet been applied to the cost estimation problem.

@@

The following chapter, Chapter 4, describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios. Further applications were also considered.

@@

* Chapter 5: Variable Selection
* Chapter 6: Regression Models
* Chapter 7: Binary Classification Models
* Chapter 8: Model blending,
* Chapter 9: Extended Analysis including overall profit analysis, research limitations and future work

The results section in each chapter presents key charts and tables that progressively answer the research question: can a statistical or machine learning model based on a company's internal historical project data predict the profitability of a new project and have a positive impact on the bottom line of the business? Then, a discussion finalises each chapter, where the results are examined deeply and analysed in a broader context. The nature of the results are also debated in terms of how surprising they were, their impact on the research hypotheses and limitations as well as suggested future work. The final chapter, the Conclusion, summarises the findings of the project, how the research answered the hypotheses and how the work could be viably applied in industry.

# Chapter 3: Literature Review

@@

#### Detailed Analysis in the Construction Industry

@@

#### Statistical and Machine Learning Models in the Construction Industry

@@

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. They found that artificial Neural Networks and SVM's were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, Neural Networks and SVM's recieved a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@Shin2015 pioneered the application of Boosted Trees to cost estimation in construction projects. This is surprisingly late given the rapid uptake of Boosted Trees over the past decade. In the study, data from 234 school building construction projects in Korea were used. Boosted trees were compared to Neural Networks, which are one of the most popular algorithms to apply to this problem. Boosted trees were found to predict costs slightly more accurately than Neural Networks, but not statstically significantly (p value < 0.05). Even though the two algorithms performed similarly, the output from Boosted Tree models provide insight into the structure of the model such as variable importance and variable partial dependence plots. This additional insight may mean Boosted Trees are preferable over Neural Networks because of their potential to engage decision makers.

A notable amount of literature has studied the predictive power of Multiple Linear Regression in the construction cost estimation problem. Often, Linear Regression is the only model assessed, without comparison to other methods such as Neural Networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on Linear Regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Out of those that have compared regression to Neural Networks, the studies have shown that Neural Networks outperform regression, however other studies established they are approximately equal [@Kim2004][@Attalla2003]. @Dissanayaka1999 found that Neural Networks outperformed multiple Linear Regression models, however regression models could be performed first to determine variable importance and condense the number of variables. It can be argued that Boosted Trees could perform as well as Neural Networks while providing additional insights (even more so than regression models) such as variable importance and partial dependence of variables against the independent variable, cost.

Despite Neural Networks typically outperforming Linear Regression models, it is interesting to note some of the outcomes of regression studies in cost estimation. Different studies tended to produce a wide range of variables that contribute at fluctuating levels in the models. This is likely because construction projects are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which contribute varying levels of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs. private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionaires, the available variables also largely depended on the questions in the questionaires, and therefore each study produced a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally reported their mean absolute error or a similar metric from the tested models. Similar result structures are useful, however there is a gap in the literature in how to translate the improved errors from the model into a well justified argument for industry to reduce losses by integrating a parametric model into their business.

@@

#### Case Based Reasoning in the Construction Industry

@@

#### Summary of Cost Estimation in the Construction Industry
